# USDA Diary & Interview File Processing

### Overview

In this brief quarto document, we are processing the USDA diary and interview files to prepare them for the project. The first step is to identify which quarter a file belongs to by its third numerical digit. After identifying the quarter it belongs to, a quarter number is assigned in a new column. Following the addition of the quarter column, the code identifies which year the file belongs to by the first two numerical digits in the file name. For example, if the file name is expd193 the file is from 2019. The year number is assigned to the file by created a new column and assigning its relevant number to it. After the years are assigned to a file, the files are binded together into a single CSV file.

## Diary Files

```{r}
library(tidyverse)
library(stringr)

# Set working directory to diary files folder
setwd("C:/Users/omyue/OneDrive/Desktop/DATA 205/USDA Expenditure Datasets/diary")

# Get list of diary files
diary_files <- list.files(pattern = "expd\\d{3}\\.csv$")

# Process and combine files
combined_diary <- diary_files %>% 
  map_dfr(~ {
    # Extract year and quarter from filename
    file_code <- str_extract(.x, "\\d{3}(?=\\.csv)")
    year <- 2000 + as.numeric(str_sub(file_code, 1, 2))
    quarter <- as.numeric(str_sub(file_code, 3, 3))
    
    # Read data - ONLY NEWID, UCC, COST
    read_csv(
      .x,
      col_types = cols(
        NEWID = col_character(),
        UCC = col_character(),
        COST = col_double(),
        .default = col_skip()  # Murder all other columns
      )
    ) %>% 
      mutate(year = year, quarter = quarter)
  })

# Save merged data
write_csv(combined_diary, "expd_merged.csv")

# Verify output
head(combined_diary)
```

## Interview Files

```{r}
library(tidyverse)
library(stringr)

# Set working directory to interview files folder
setwd("C:/Users/omyue/OneDrive/Desktop/DATA 205/USDA Expenditure Datasets/interview")

# Get list of interview files (supports Q1 "x" suffix)
interview_files <- list.files(pattern = "fmli\\d{3}.*\\.csv$")

# Process and combine files
combined_interview <- interview_files %>% 
  map_dfr(~ {
    # Extract 3-digit code after "fmli" (ignores "x" in Q1 filenames)
    file_code <- str_extract(.x, "(?<=fmli)\\d{3}")  # Regex magic
    
    # Extract year and quarter
    year <- 2000 + as.numeric(str_sub(file_code, 1, 2))
    quarter <- as.numeric(str_sub(file_code, 3, 3))
    
    # Read data - ONLY your specified columns
    read_csv(
      .x,
      col_types = cols(
        NEWID = col_character(),
        FINCBTTXM = col_double(),    # Total income
        INCLASS = col_character(),   # Income class
        HRSIZE = col_integer(),      # Household size
        REGION = col_character(),    # Region
        AGE_REF = col_integer(),     # Age of reference person
        .default = col_skip()        # Destroy other columns
      )
    ) %>% 
      mutate(year = year, quarter = quarter)
  })

# Save merged data
write_csv(combined_interview, "fmli_merged.csv")

# Verify output
head(combined_interview)
```
